---
title: "Fundamentals of Data Wrangling and Analysis Examples"
author: "R for the Rest of Us"
output: 
    html_document:
        css: slides/style.css
        toc: true
        toc_float: true
        df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(kableExtra)
```

# Import Data

```{r}
nhanes <- read_csv("data/nhanes.csv") %>%
  clean_names()
```


# `select`

```{r}
nhanes %>%
  select(age)
```




## Other ways we can `select` variables

Select multiple variables by listing them:

```{r}
nhanes %>%
  select(height, weight)
```



## Other ways we can `select` variables

You can use `one_of()` to do the same thing:

```{r}
nhanes %>%
  select(one_of(c("height", "weight")))
```



## Other ways we can `select` variables

`contain()` selects variable with certain text in the variable name:


```{r}
nhanes %>%
  select(contains("age"))
```



## Other ways we can `select` variables

You can select a range of columns using the var1:var2 pattern


```{r}
nhanes %>%
  select(weight:bmi)
```



## Other ways we can `select` variables

Drop variables using the -var format:


```{r}
nhanes %>%
  select(-id)
```



## Other ways we can `select` variables

Drop a set of variables using the -(var1:var2) format:


```{r}
nhanes %>%
  select(-(id:education))
```

# `mutate`

## 1. Create a **new variable with a specific value**

```{r}
nhanes %>%
  mutate(country = "United States") %>% 
  select(country)
```




## 2. Create a **new variable based on other variables**

```{r}
nhanes %>%
  mutate(height_inches = height / 2.54) %>% 
  select(contains("height"))
```






## 3. Change an **existing variable**

```{r}
nhanes %>%
  mutate(bmi = round(bmi, digits = 1)) %>% 
  select(bmi)
```



# `filter`


## `filter`

```{r}
nhanes %>% 
  filter(gender == "female") %>%
  select(gender)
```






## `filter`

```{r}
nhanes %>% 
  filter(health_gen != "Good") %>%
  select(health_gen)
```



## `filter`

```{r}
nhanes %>% 
  filter(health_gen != "Good" | health_gen != "VGood" | health_gen != "Excellent") %>%
  select(health_gen)
```




## `filter`

```{r}
nhanes %>% 
  filter(health_gen %in% c("Good", "VGood", "Excellent")) %>%
  select(health_gen)
```




## `filter`

You can chain together multiple `filter` functions. Doing it this way, we don't have create complex logic in one line.

```{r}
nhanes %>% 
  filter(gender == "male") %>%
  filter(health_gen %in% c("Good", "VGood", "Excellent")) %>%
  select(gender, health_gen)
```





## `filter`

```{r}
nhanes %>% 
  filter(age > 50)
```




## `filter`

You can drop `NAs` with `!is.na()` 

```{r}
nhanes %>% 
  filter(age > 50) %>% 
  filter(!is.na(marital_status)) %>%
  select(age, marital_status)
```





## `filter`

You can also drop `NAs` with `drop_na()`

```{r}
nhanes %>% 
  filter(age > 50) %>% 
  drop_na(marital_status) %>%
  select(age, marital_status)
```



# `summarize`

This doesn't work! Notice what the result is.

```{r}
nhanes %>% 
  summarize(mean_active_days = mean(phys_active_days))
```





## `summarize`

Add na.rm = TRUE to make this work.


```{r}
nhanes %>% 
  summarize(mean_active_days = mean(phys_active_days,
                                    na.rm = TRUE))
```




## `summarize`


You can have multiple arguments in each usage of `summarize`.

```{r}
nhanes %>% 
  summarize(mean_active_days = mean(phys_active_days, na.rm = TRUE),
            median_active_days = median(phys_active_days, na.rm = TRUE),
            number_of_responses = n())
```




# `group_by`


```{r}
nhanes %>% 
  group_by(age_decade) %>%
  summarize(mean_active_days = mean(phys_active_days,
                                    na.rm = TRUE)) 
```







## `group_by` example

`group_by` works for multiple groups.

```{r}
nhanes %>% 
  group_by(age_decade, gender) %>%
  summarize(mean_active_days = mean(phys_active_days,
                                    na.rm = TRUE)) 
```




## `count`

If you just want to count the number of things per group, you can use `count`.


```{r}
nhanes %>% 
  count(age_decade)
```




## `count`

You can also count by multiple groups.

```{r}
nhanes %>% 
  count(age_decade, gender)
```





# `arrange`



## `arrange` example

R arranges in ascending order by default

```{r}
nhanes %>% 
  group_by(age_decade, gender) %>%
  summarize(mean_active_days = mean(phys_active_days,
                                    na.rm = TRUE)) %>% 
  arrange(mean_active_days) 
```



## `arrange` example

You can also arrange in descending order


```{r}
nhanes %>% 
  group_by(age_decade, gender) %>%
  summarize(mean_active_days = mean(phys_active_days,
                                    na.rm = TRUE)) %>% 
  arrange(desc(mean_active_days)) 
```



# Crosstabs

Sometimes you want your results in a crosstab. We'll use the `tabyl` function in `janitor` package to make crosstabs automatically.


```{r}
nhanes %>% 
  tabyl(gender, age_decade) 
```



## `adorn_` functions

`janitor` has a set of functions that all start with `adorn_` that add a number of things to our crosstabs. You call them after `tabyl`.


## Add totals

```{r}

nhanes %>% 
  tabyl(gender, age_decade) %>% 
  adorn_totals(c("row", "col"))
  

```





## Add percentages


```{r}
nhanes %>% 
  tabyl(gender, age_decade) %>% 
  adorn_totals(c("row", "col")) %>% 
  adorn_percentages()
```



## Format percentages

```{r}
nhanes %>% 
  tabyl(gender, age_decade) %>% 
  adorn_totals(c("row", "col")) %>% 
  adorn_percentages() %>% 
  adorn_pct_formatting() 
```





## Include n alongside percentages

```{r}
nhanes %>% 
  tabyl(gender, age_decade) %>% 
  adorn_totals(c("row", "col")) %>% 
  adorn_percentages() %>% 
  adorn_pct_formatting() %>% 
  adorn_ns()
```




## Add title

```{r}
nhanes %>% 
  tabyl(gender, age_decade) %>% 
  adorn_totals(c("row", "col")) %>% 
  adorn_percentages() %>% 
  adorn_pct_formatting() %>% 
  adorn_ns() %>% 
  adorn_title()
```






## Three (or more) way crosstabs

You can also do three (or more) way crosstabs automatically by adding more variables to the `tabyl` function.

```{r}
nhanes %>% 
  tabyl(gender, age_decade, education) %>%
  adorn_totals(c("row", "col")) %>% 
  adorn_percentages() %>% 
  adorn_pct_formatting() %>% 
  adorn_ns() %>% 
  adorn_title() 

```

# Create new data frames

Sometimes you want to save the results of your work to a new data frame. 

```{r}
phys_activity_by_age <- nhanes %>% 
  group_by(age_decade) %>% #<<
  summarize(mean_active_days = mean(phys_active_days,
                                    na.rm = TRUE)) %>% 
  drop_na()
```

```{r}
phys_activity_by_age
```

